Open-source tools in R for landscape ecology

### 2021-01-26 paper/paper.Rmd
update abstract, ljg affiliation, glossary, and minor typos

---
title: Open-source tools in R for landscape ecology
# subtitle: Do you have a subtitle? If so, write it here
# titlerunning: Open source landscape ecology tools
# authorrunning: Short form of author list if too long for running head
# thanks: |
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Büsgenweg 3, 37077 Göttingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- R packages
- reproducibility
- scientific software
- ...
abstract: |
  max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

API & Application programming interface \\
cell & Smallest, rectangular unit of raster data model \\
CRAN & Comprehensive R Archive Network \\
CRS & Coordinate references systems \\
GIS & Geographical information system \\
landscape & Mosaic of land covers, ecosystems, habitat types, land uses \\
NLM & Neutral landscape model \\
open-source software & Software released under licenses that allow to use, modify and distribute it \\
patch & Neighboring cells with same value \\
raster & Cell grid based data model \\
resolution & Size in meter or degree of one cell \\
R package & User-created software extension to R programming language \\
SDM & Species distribution modeling \\
vector & Points, lines, and polygons-based data model \\

@@

While *R* has several built-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
Importantly, main *R* packages for spatial data use the external *GDAL* \cite{GDAL} and *PROJ* \cite{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, *R* allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs.
Popular approaches and packages include generalized linear models using, e.g., the *stats* \cite{RCoreTeam2019} package; generalized additive models using, e.g., the *mgcv* \cite{Wood2017} or *lme4* \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the *rpart* \cite{Therneau2019}, *randomForest* \cite{Liaw2002} or *ranger* \cite{Wright2017} package or multivariate data analysis using , e.g., the *ade4* \cite{Dray2007} or *vegan* \cite{Oksanen2019} package.
Also, packages specifically designed for SDM exists, including includes the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020}, *PresenceAbsence* \cite{Freeman2008}, or *zoon* \cite{Golding2018} packages.
<!-- These packages [check!] allow the user to create ensemble models of SDMs based on different underlying model structures such as Maxent, regression trees, and generalized linear or additive models.
This is considered the gold standard in contemporary SDM \cite(Araujo2007).-->

@@

Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:connectivity}) are also important for landscape genetics.
Further functionality for landscape genetics such as ... <!-- LJG: I think a little more info needed here - happy to dig through the packages to add in, but thought I'd leave for now incase anyone knows off top of their heads -->can be found in the *graphs4lg* \cite{Savary2020}, *PopGenReport* \cite{Adamack2014,Gruber2015}, *HierDpart* packages \cite{Qin2019}, or *GeNetIt* \cite{Murphy2010}.

@@

@@

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey) B).
Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").

@@

Interestingly, more people use the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%).
This was also represented in the most used *R* packages (Fig. \@ref(fig:fig-survey) C)).
When asked for the three most used packages, participants of the survey listed `r length(unique(r_packages))` packages in total.
The *raster* package was mentioned the most, followed by the *sf* package.
Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \@ref(fig:fig-survey) C).
Nevertheless, the large availability and usage of different *R* packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; `r round(r_packages_df[[1, 2]], 2)`%).

@@

<!--a paragraph summarizing possibilities-->
A comprehensive collection of *R* software package exist to handle the most common tasks and fields of landscape ecology. 
Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, *R* allows for transparent and reproducible workflows.
Additionally, this also allows to easily interchange, modify, or adapt methods from other related and unrelated fields.

<!--a paragraph summarizing the survey-->
The survey revealed that overall the landscape ecology community is well pleased with the capabilities of the *R* programming language for landscape ecology. 
Furthermore, the survey showed the highly dynamic development of *R* packages by the community itself, showing that avaiable tools are constantly updated.

>>>>

---
title: Open-source tools in R for landscape ecology
# subtitle: Do you have a subtitle? If so, write it here
# titlerunning: Open source landscape ecology tools
# authorrunning: Short form of author list if too long for running head
# thanks: |
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Büsgenweg 3, 37077 Göttingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- R packages
- reproducibility
- scientific software
- ...
abstract: |
  Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology. In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of *R* users within the landscape ecology community. In this we find that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually updating the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

API & Application programming interface. A set of protocols for sending and retrieving information from a server \\
cell & Smallest, rectangular unit of raster data model \\
CRAN & Comprehensive R Archive Network. A repository for R packages \\
CRS & Coordinate references systems. A coordinate-based local, regional or global system used to locate geographical entities \\
GIS & Geographical information system \\
landscape & Mosaic of land covers, ecosystems, habitat types, land uses \\
NLM & Neutral landscape model \\
open-source software & Software released under licenses that allow to use, modify and distribute it \\
patch & Neighboring cells with same value \\
raster & Cell grid based data model \\
resolution & Size in meter or degree of one cell \\
R package & User-created software extension to R programming language \\
SDM & Species distribution modelling \\
vector & Points, lines, and polygons-based data model \\

@@

While *R* has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
Importantly, main *R* packages for spatial data use the external *GDAL* \cite{GDAL} and *PROJ* \cite{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, *R* allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs.
Popular approaches and packages include generalized linear models using, e.g., the *stats* \cite{RCoreTeam2019} package; generalized additive models using, e.g., the *mgcv* \cite{Wood2017} or *lme4* \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the *rpart* \cite{Therneau2019}, *randomForest* \cite{Liaw2002} or *ranger* \cite{Wright2017} package or multivariate data analysis using , e.g., the *ade4* \cite{Dray2007} or *vegan* \cite{Oksanen2019} package.
Also, packages specifically designed for SDM exist, including the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020}, *PresenceAbsence* \cite{Freeman2008}, or *zoon* \cite{Golding2018} packages. Additionally, packages such as *ENMeval* \cite{R-enmeval} provide functionality for model tuning and evaluation. 
<!-- These packages [check!] allow the user to create ensemble models of SDMs based on different underlying model structures such as Maxent, regression trees, and generalized linear or additive models.
This is considered the gold standard in contemporary SDM \cite(Araujo2007).-->

@@

Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:connectivity}) are also important for landscape genetics.
Further functionality for landscape genetics such as quantifying and analysing population genetic structure, and hierarchical decomposition analysis can be found in the *graphs4lg* \cite{Savary2020}, *PopGenReport* \cite{Adamack2014,Gruber2015}, *HierDpart* packages \cite{Qin2019}, or *GeNetIt* \cite{Murphy2010}.

@@

& ENMeval & Model tuning and evaluation & \cite{R-enmeval} \\

@@

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey) B).
The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").

@@

More people use the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%).
This was also represented in the most used *R* packages (Fig. \@ref(fig:fig-survey) C)).
When asked for the three most used packages, participants of the survey listed `r length(unique(r_packages))` packages in total.
The *raster* package was mentioned the most, followed by the *sf* package.
Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \@ref(fig:fig-survey) C).
Nevertheless, the large availability and usage of different *R* packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; `r round(r_packages_df[[1, 2]], 2)`%).

@@

<!--a paragraph summarizing possibilities-->
A comprehensive collection of *R* software packages exist to handle the most common tasks and fields of landscape ecology. 
Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, *R* allows for transparent and reproducible workflows.
Additionally, this also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.

<!--a paragraph summarizing the survey-->
The survey revealed that overall the landscape ecology community is satisfied with the capabilities of the *R* programming language for landscape ecology. 
Furthermore, the survey showed the highly dynamic development of *R* packages by the community itself, showing that avaiable tools are constantly updated.

### 2021-01-26 paper/paper.tex
update abstract, ljg affiliation, glossary, and minor typos

\institute{
        Maximillian H.K. Hesselbarth \at
     a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA \\
     \email{\href{mailto:mhk.hesselbarth@gmail.com}{\nolinkurl{mhk.hesselbarth@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Jakub Nowosad \at
     Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland \\
     \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Johannes Signer \at
     Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Büsgenweg 3, 37077 Göttingen, Germany \\
     \email{\href{mailto:jsigner@uni-goettingen.de}{\nolinkurl{jsigner@uni-goettingen.de}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Laura J. Graham \at
     Geography, Earth \& Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom \\
     \email{\href{mailto:l.graham@bham.ac.uk}{\nolinkurl{l.graham@bham.ac.uk}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

\begin{abstract}
max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
\\
\keywords{
        spatial data \and
        statistical programming language \and
        R packages \and
        reproducibility \and
        scientific software \and
        \ldots{} \and
    }

@@

While \emph{R} has several built-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \emph{R} packages (\cite{Lovelace2019}, page 10, Table \ref{tab:packages}).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \emph{R} language.
Importantly, main \emph{R} packages for spatial data use the external \emph{GDAL} \cite{GDAL} and \emph{PROJ} \cite{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, \emph{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of \emph{R} packages used for SDMs.
Popular approaches and packages include generalized linear models using, e.g., the \emph{stats} \cite{RCoreTeam2019} package; generalized additive models using, e.g., the \emph{mgcv} \cite{Wood2017} or \emph{lme4} \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the \emph{rpart} \cite{Therneau2019}, \emph{randomForest} \cite{Liaw2002} or \emph{ranger} \cite{Wright2017} package or multivariate data analysis using , e.g., the \emph{ade4} \cite{Dray2007} or \emph{vegan} \cite{Oksanen2019} package.
Also, packages specifically designed for SDM exists, including includes the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020}, \emph{PresenceAbsence} \cite{Freeman2008}, or \emph{zoon} \cite{Golding2018} packages.

@@

Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:connectivity}) are also important for landscape genetics.
Further functionality for landscape genetics such as \ldots{} can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, \emph{HierDpart} packages \cite{Qin2019}, or \emph{GeNetIt} \cite{Murphy2010}.

@@

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').

Interestingly, more people use the raster data model (72.8\%) in comparison to the vector data model (27.2\%).
This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-survey} C)).
When asked for the three most used packages, participants of the survey listed 83 packages in total.
The \emph{raster} package was mentioned the most, followed by the \emph{sf} package.
Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \ref{fig:fig-survey} C).
Nevertheless, the large availability and usage of different \emph{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 33.44\%).

@@

A comprehensive collection of \emph{R} software package exist to handle the most common tasks and fields of landscape ecology.
Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, \emph{R} allows for transparent and reproducible workflows.
Additionally, this also allows to easily interchange, modify, or adapt methods from other related and unrelated fields.

The survey revealed that overall the landscape ecology community is well pleased with the capabilities of the \emph{R} programming language for landscape ecology.
Furthermore, the survey showed the highly dynamic development of \emph{R} packages by the community itself, showing that avaiable tools are constantly updated.

>>>>

\institute{
        Maximillian H.K. Hesselbarth \at
     a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA \\
     \email{\href{mailto:mhk.hesselbarth@gmail.com}{\nolinkurl{mhk.hesselbarth@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Jakub Nowosad \at
     Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland \\
     \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Johannes Signer \at
     Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Büsgenweg 3, 37077 Göttingen, Germany \\
     \email{\href{mailto:jsigner@uni-goettingen.de}{\nolinkurl{jsigner@uni-goettingen.de}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Laura J. Graham \at
     a) Geography, Earth \& Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom b) Biodiversity, Ecology \& Conservation Group, IIASA, Vienna, Austria \\
     \email{\href{mailto:l.graham@bham.ac.uk}{\nolinkurl{l.graham@bham.ac.uk}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

\begin{abstract}
Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the \emph{R} programming language provides a wealth of community developed tools for landscape ecology. In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of \emph{R} users within the landscape ecology community. In this we find that landscape ecologists are generally satisfied with the functionality available within \emph{R}, and that as a community they are continually updating the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools.
\\
\keywords{
        spatial data \and
        statistical programming language \and
        R packages \and
        reproducibility \and
        scientific software \and
        \ldots{} \and
    }

@@

While \emph{R} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \emph{R} packages (\cite{Lovelace2019}, page 10, Table \ref{tab:packages}).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \emph{R} language.
Importantly, main \emph{R} packages for spatial data use the external \emph{GDAL} \cite{GDAL} and \emph{PROJ} \cite{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, \emph{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of \emph{R} packages used for SDMs.
Popular approaches and packages include generalized linear models using, e.g., the \emph{stats} \cite{RCoreTeam2019} package; generalized additive models using, e.g., the \emph{mgcv} \cite{Wood2017} or \emph{lme4} \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the \emph{rpart} \cite{Therneau2019}, \emph{randomForest} \cite{Liaw2002} or \emph{ranger} \cite{Wright2017} package or multivariate data analysis using , e.g., the \emph{ade4} \cite{Dray2007} or \emph{vegan} \cite{Oksanen2019} package.
Also, packages specifically designed for SDM exist, including the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020}, \emph{PresenceAbsence} \cite{Freeman2008}, or \emph{zoon} \cite{Golding2018} packages. Additionally, packages such as \emph{ENMeval} \cite{R-enmeval} provide functionality for model tuning and evaluation.

@@

Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:connectivity}) are also important for landscape genetics.
Further functionality for landscape genetics such as quantifying and analysing population genetic structure, and hierarchical decomposition analysis can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, \emph{HierDpart} packages \cite{Qin2019}, or \emph{GeNetIt} \cite{Murphy2010}.

@@

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').

More people use the raster data model (72.8\%) in comparison to the vector data model (27.2\%).
This was also represented in the most used \emph{R} packages (Fig. \ref{fig:fig-survey} C)).
When asked for the three most used packages, participants of the survey listed 83 packages in total.
The \emph{raster} package was mentioned the most, followed by the \emph{sf} package.
Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \ref{fig:fig-survey} C).
Nevertheless, the large availability and usage of different \emph{R} packages across the community can be seen in the large ``others'' option (packages mentioned by less than 5 participants; 33.44\%).

@@

A comprehensive collection of \emph{R} software packages exist to handle the most common tasks and fields of landscape ecology.
Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, \emph{R} allows for transparent and reproducible workflows.
Additionally, this also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.

The survey revealed that overall the landscape ecology community is satisfied with the capabilities of the \emph{R} programming language for landscape ecology.
Furthermore, the survey showed the highly dynamic development of \emph{R} packages by the community itself, showing that avaiable tools are constantly updated.
